---
title: "Assignment 2"
author: "Jasper Sänger"
date: "2024-12-3"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup & libraries, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = F,
  message = F,
  fig.align = "center", 
  fig.width = 10,
  fig.height = 4
)

#loading necessary libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(tsibble)
library(fable)
library(feasts)
library(lubridate)
library(patchwork)
library(gridExtra)
```

```{r same theme throughout for ggplot}
theme_set(theme_bw())
```

## I. Assessing Trend

*•* Assess the need for a trend component and include it in the model,
as you see fit. I*f needed, reduce the data to model only those
observations for which the trend is consistent. Note that, if you decide
to reduce the data, to be consistent from here on you must work with the
reduced data.*

### Series 1

```{r given by teacher}
data1.tibble <- readr::read_csv("assignment_1-2024-data1.csv", show_col_types = F)
colnames(data1.tibble) <- c('t', 'value')
data1 <- data1.tibble %>%
mutate(t = yearmonth(t)) %>%
as_tsibble(index = t)
ggplot() +
geom_line(aes(x = t, y = value), data = data1 %>% slice(1:56), lty = 2, col = 'red2') +
geom_line(aes(x = t, y = value), data = data1 %>% slice(57:n()))
```

For `data1`, an upwards trend is visible from around January 2015
onward. For further analysis, exploration and modelling, the data set is
modified and will no longer consider data prior to year 2015. Below, the
condensed data1_final graph is displayed, alongside a trend line,
indicating this upward trend.

```{r reducing data1}
data1_final <- data1 %>% filter(t >= yearmonth("2015 Jan"))
```

```{r cleaning environment}
rm(list = setdiff(ls(), c("data1_final", "data2")))
```

```{r plotting data1_final with trend line}
data1_final %>% ggplot(aes(x = t, y = value)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Final Data 1 with Trend Line (Jan 2015 onwards)", x = "Time", y = "Value", linetype = "Trend") +
  theme_minimal()
```

```{r data1 residual analysis}
data1_trend_model <- data1_final %>% model(TSLM(value ~ trend()))

gg_tsresiduals(data1_trend_model) +
  labs(title = "Residual Diagnostics for Data 1 Trend Model",
       x = "Time", y = "Residuals")
```

Starting with the trend model residuals, the model displays that the
distance between fitted values seem to average out to 0 and do not
follow a clear pattern, with variance increasing from 2022 to 2024.
Based on this graph alone, one can conclude that the model fits the data
quite well. Notably, the increasing variance towards later years
potentially indicates heteroskedasticity, which the model fails to
capture in that period.

The ACF of the residuals displays two significant values at lag 8 and 9,
but no further pattern. This suggests that the model has room for
improvement as there may still be some unexplained structure in the
data. However, a broader pattern cannot be identified indicating that
the model captures most of the systematic components of the data.

The distribution of residuals follows an almost normal distribution
(bell shaped), a characteristic of a good model.

### Series 2

```{r plotting data_2 with trend line}
data2 <- readr::read_csv("assignment_1-2024-data2.csv", show_col_types = F) %>% 
  as_tsibble(index = t)

data2 %>% 
  autoplot(u) +
  geom_smooth(method = lm, formula = y ~ x, se = FALSE)
```

A slight downward trend can be spotted for `data2`. Subsequent residual
analysis is necessary to explore the need and impact of incorporating
trend in a model.

```{r data2 residual analysis}
data2_trend_model <- data2 %>% model(TSLM(u ~ trend()))
data2_trend_model.augmented <- data2_trend_model%>%
  augment()

gg_tsresiduals(data2_trend_model)+
    labs(title = "Residual Diagnostics for Data 2 Trend Model",
       x = "Time", y = "Residuals")

```

The residuals for `data2`are dispersed around zero but seem to follow a
systematic pattern over time.

The ACF of the residuals displays significant values at each of the 30
lags, with a clear seasonality pattern. With spikes every 7 lags, this
could indicate weekly seasonality. This means that the trend model may
be missing a seasonal component and could benefit from incorporating
one.

The histogram of the residuals is bell shaped and normally distributed
around 0, which is a positive aspect of the model. However, the other
two residual plots indicate that the trend model would require a
seasonal component to improve.

```{r report of trend model series 2}
data2_trend_model%>%
  report()
```

Although the coefficients are statistically significant (at the 1%
significance level) for this model, they are extremely small, so is the
R-squared value. The trend component does not improve the model's
explanatory power by much and is therefore not considered for further
modelling of `data2`.

## II. Assessing Seasonality

*• Check the need for seasonal fixed effects in the residuals of the
model(s) produced in the previous point.*

• Multiple seasonal fixed effects might be needed. Check them in the
order discussed in class.

### Series 1

```{r creating monthly dummies model}
data1_monthly_dummies = data1_final %>% mutate(month = factor(month(t)))

data1_monthly_dummies_trend <- data1_monthly_dummies%>%
  model(TSLM(value ~ trend() + month))
```

```{r residuals plot of monthly dummies model}
gg_tsresiduals(data1_monthly_dummies_trend) +
labs(title = "Residual Diagnostics for Data 1 Monthly (+Trend) Model",
       x = "Time", y = "Residuals")
```

The residuals of the trend model incorporating monthly seasonality
display residuals mostly centered around 0 in the time series of the
residuals as well as the histogram. The ACF has two significant lags (8
and 9) and does not show any further pattern. This is comparable to the
residual analysis for the trend model without incorporating seasonality.

```{r report of trend & seasonality model series 1}
data1_monthly_dummies_trend %>% 
  report()
```

The coefficients for month 1 (intercept after controlling trend) and
each of the monthly dummies are not statistically significant. Only
month 11, which has a value of 15.02 higher that the baseline while
controlling for trend, is significant at the 10% level and could be
considered to be incorporated. Although the R-squared is high, this is
almost solely to the highly significant `trend()` component of the
model. The lack of statistical significance for most dummies suggests
that monthly seasonality might not be the best way to model periodic
effects.

### Series 2

```{r creating seasonality dummies model, starting with larger seasonality}
data2_dummies <- data2 %>%
  mutate(
    month = factor(month(t)),          
    day_of_week = factor(wday(t))
  )

data2_dummies_trend_month <- data2_dummies %>%
  model(TSLM(u ~ trend() + month))

data2_dummies_trend_day <- data2_dummies %>%
  model(TSLM(u ~ trend() + day_of_week))

data2_dummies_trend_month_day <- data2_dummies %>%
  model(TSLM(u ~ trend() + month + day_of_week))
```

```{r monthly seasonality residual analysis}
gg_tsresiduals(data2_dummies_trend_month) +
  labs(title = "Residual Diagnostics for Data 2 Monthly (+Trend) Model",
       x = "Time", y = "Residuals")

```

The residual analysis shows similar visualizations as for the trend
model without monthly dummies. Just as before, this suggests that the
model is still not capturing all seasonal components present in `data2`.

```{r report of trend & monthly seasonality model series 2}
data2_dummies_trend_month%>% 
  report()
```

The coefficients of several monthly dummies are significant at the 1%
level, while `month7` is significant at the 5% level and `month2` and
`month11` at the 10% level. The R-squared increased by more than 10%
points by adding monthly dummies, highlighting that monthly seasonality
dummies has a substantial impact on how much variance in the data is
explained.

```{r day of week dummies residual analysis}
gg_tsresiduals(data2_dummies_trend_day) +
  labs(title = "Residual Diagnostics for Data 2 Day-of-Week (+Trend) Model",
       x = "Time", y = "Residuals")
```

This residual analysis for the trend model that incorporates day of week
dummies shows that the residuals are centered around 0 but exhibit a
clear pattern, indicating that some structure remains unexplained
(potentially the monthly seasonality from before or something else). The
ACF plot reveals significant autocorrelations across multiple lags,
suggesting the model does not fully account for temporal dependencies.
The histogram of residuals is approximately bell-shaped, indicating
near-normality but with potential room for improvement in the model's
fit.

```{r report of trend & day of week model series 2}
data2_dummies_trend_day%>% 
  report()
```

The model shows significant coefficients for all week day dummies. The
R-squared has also increased by approximately 12 percentage points
compared to the model which only captures trend.

```{r 2 levels of seasonality residual analysis}
gg_tsresiduals(data2_dummies_trend_month_day)+
    labs(title = "Residual Diagnostics for Data 2 Two Levels of Seasonality (+Trend) Model",
       x = "Time", y = "Residuals")
```

The residual diagnostics for the combined levels of seasonality dummies
model indicate improvements compared to individual models. However, the
residuals still show structured patterns and significant
autocorrelations remain in the ACF. This suggests further adjustments
may be needed to fully capture dependencies based on time.

```{r report of trend & 2 levels of seasonality series 2}
data2_dummies_trend_month_day%>% 
  report()
```

The model combining trend, month, and day dummies shows multiple
statistically significant coefficients for both seasonal and trend
components. The R-squared reflects better explanatory power compared to
simpler models, but there is still room for improvement (as also
indicated in the residual analysis).

## III. Include Events

*• Including events - Christmas day, as well as for the day before and
the day after.*

```{r Christmas dummy variables}
data2 %>% 
  mutate(
    month = factor(month(t)), 
    day_of_week = factor(wday(t)), 
    christmas = day(t) == 25 & month(t) == 12,
    day_before_christmas = day(t) == 24 & month(t) == 12,
    day_after_christmas = day(t) == 26 & month(t) == 12
  ) %>% 
  relocate(t, u, christmas, day_before_christmas, day_after_christmas) -> data2_christmas

data2_model_christmas <- data2_christmas %>%
  model(TSLM(u ~ trend() + month + day_of_week + christmas + day_before_christmas + day_after_christmas))

data2_model_christmas%>%
  report()
```

None of the coefficients of the special day dummies (Christmas, day
before and after) incorporated in this model are statistically
significant. Therefore, their inclusion does not improve the model.

## IV Predictions & accuracy

*•* Split the data into a training and a testing set, make predictions,
visualize, and compute accuracy metrics.

*• Do this in the same way as in Assignment 1: follow those
instructions.*

*• Include the benchmark models (only in this part, no need to include
them during model building).*

*• If you have reduced any of the data and the split is a percentage,
you can apply the percentage to the reduced data, rounding as needed.*

```{r Splitting data & predictions}
data2_dummies <- data2 %>%
  mutate(
    month = factor(month(t)),          
    day_of_week = factor(wday(t))
  )

split_index <- round(0.8 * nrow(data1_final))
train_data1 <- data1_final %>% filter(row_number() <= split_index)  
test_data1 <- data1_final %>% filter(row_number() > split_index)

cutoff_date <- as.Date("2023-02-01")  
test_data2 <- data2_dummies %>% filter(t >= cutoff_date)
train_data2 <- data2_dummies %>% filter(t < cutoff_date)

# All-in-One Forecast Models
# Data1
horizon_data1 <- nrow(test_data1)
forecasts_all_data_1 <- train_data1 %>%
  model(
    Naive = NAIVE(value),
    Mean = MEAN(value),
    Snaive = SNAIVE(value),
    Drift = RW(value ~ drift()),
    Trend = TSLM(value ~ trend()) # New Model
  ) %>%
  forecast(h = horizon_data1)

# Data2
horizon_data2 <- nrow(test_data2)
forecasts_all_data2 <- train_data2 %>%
  model(
    Naive = NAIVE(u),
    Mean = MEAN(u),
    Snaive = SNAIVE(u),
    Drift = RW(u ~ drift()),
    TSLM(u ~ trend() + month + day_of_week)
  ) %>%
  forecast(new_data = test_data2)

```

```{r Visualizing models and predictions}
Model_data1 <-train_data1 %>%
  model(TSLM(value ~ trend()))%>%
  forecast(h = horizon_data1)

Model_data2 <-train_data2%>%
  model(TSLM(u ~ trend() + month + day_of_week))%>%
  forecast(new_data = test_data2)

grid.arrange(
  autoplot(Model_data1, train_data1) +
    autolayer(test_data1, colour = "red", size = 0.4, linetype = "dashed") +
    labs(
      title = "Models Data 1: Forecast and Observations",
      x = "Time",
      y = "Value"
    ) +
    theme_minimal(),
  
  autoplot(Model_data2, train_data2) +
    autolayer(test_data2, colour = "red", size = 0.4, linetype = "dashed") +
    labs(
      title = "Models Data 2: Forecast and Observations",
      x = "Time",
      y = "Value"
    ) +
    theme_minimal(),
  ncol = 1 
)
```

```{r Accuracy metrics for data1 and data2}
accuracy_metrics1 <- forecasts_all_data_1 %>%
  accuracy(test_data1)

accuracy_metrics2 <- forecasts_all_data2 %>%
  accuracy(test_data2)
```

```{r Data1}
accuracy_metrics1
```

The Trend model achieves the best accuracy, with the lowest root mean
square error (RMSE) of 25.72 and mean absolute error (MAE) of 20.18,
demonstrating its superior ability to capture the underlying trends in
`data1` compared to baseline models.

```{r Data2}
accuracy_metrics2
```

The Trend model incorporating trend, monthly, and weekday effects
achieves the highest accuracy for `data2`compared to other simple
models, with the lowest root mean square error (RMSE) of 1.02 and mean
absolute error (MAE) of 0.80.

```{r Median Absolute Deviation for Data1}
tibble(
  .model = forecasts_all_data_1$.model,          
  fc_error = forecasts_all_data_1$.mean - test_data1$value ) %>%
  group_by(.model) %>%                       
  summarise(MAD = median(abs(fc_error), na.rm = TRUE)) 
```

The Trend model demonstrates the best performance for `data1`, with a
median absolute deviation (MAD) of 17.10.

```{r Median Absolute Deviation for Data2}
tibble(
  .model = forecasts_all_data2$.model,
  fc_error = forecasts_all_data2$.mean- test_data2$u)%>%
  group_by(.model)%>%
    summarise(MAD = median(abs(fc_error), na.rm = TRUE))


```

For `data2,` the combined model with trend, monthly, and day-of-week
effects outperforms all others, achieving the lowest median absolute
deviation (MAD) of 0.60.
